### 2015-03-24 thesis-proposal/problem.tex
Reworked proposal

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Introduction}\label{motivation}
% \begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
% and generality in language design is a serious unsolved 
% problem.} -- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
%%One might exp ``generality''
Well-designed general-purpose programming languages are remarkably stable: they rarely need to be revised or forked into dialects, because programmers can express the constructs that they need just as well in orthogonal libraries. %Instead, . . %Indeed, a preponderence of new dialects is clear evidence against claims of generality.  %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
Abstraction mechanisms informed by type theory, like  the ML module system \cite{MacQueen:1984:MSM:800055.802036}, have put this ideal within reach. 
Yet despite their expressive power, new {dialects} of languages like ML continue to arise. Why is this?

% Unlike libraries organized by a module system, language dialects are not trivially composable.%Constructing a software system using libraries written in different dialects is difficult because they cannot always be composed. %This suggests that library-based mechanisms are still not general enough to encompass many desirable modes of expression.   %Library implementations are not yet satisfying.
%Why would a well-informed programmer consider forming a new dialect? 
Perhaps the most common justification is that the \emph{syntactic cost} of a desirable construct is not  preserved when it is expressed using general-purpose mechanisms. %Syntactic cost is often assessed qualitatively \cite{green1996usability} (though quantitative metrics can also be defined). 
For example, the semantics of regular expressions can be expressed using abstract data types (which in ML are a mode of use of the module system), but as we will detail in Sec. \ref{sec:syntax}, this presents some syntactic difficulties. For programmers in problem domains  where regular expressions are common, e.g. bioinformatics, it is tempting to design a ``domain-specific'' dialect that differs only in that it includes derived syntax (i.e. ``syntactic sugar'') for regular expression patterns. There is no shortage of tools that aid in the construction of such syntactic dialects (e.g. Camlp4 \cite{ocaml-manual}, SugarJ \cite{erdweg2011sugarj}, SugarHaskell \cite{erdweg2012layout} and many others).%Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).

Somewhat more rarely, dialects are motivated by the need to extend a general-purpose language with new derived semantic constructs, i.e. constructs that cannot be expressed as purely syntactic sugar, but where the general-purpose language is a suitable translation target. As a minimal example, consider System \textbf{F}, i.e. the polymorphic lambda calculus (perhaps the simplest general-purpose language) \cite{pfpl}. If we wanted to express record types in \textbf{F}, we could not do so simply with derived syntax. We could, however, develop a dialect that included the type and term operators relevant to records, specifying their static semantics directly but their dynamic semantics by translation to the polymorphic lambda calculus (using a standard Church-style encoding).\todo{couple more sentences} %Constructs specified by a type-directed translation like this can be considered \emph{derived semantic constructs}.

@@

The problem with designing new language dialects in these situations is that a program written in one dialect cannot safely and idiomatically interface with libraries written in a different dialect. %At best, one can hope that the compilers for the two languages target a common intermediate language,  
To do so would require combining the two dialects into a single language, but there is no systematic method to do so, much less one that guarantees that important metatheoretic properties established about each dialect in isolation will be conserved. As a simple example, two dialects might be known to have an unambiguous concrete syntax in isolation, but if the syntax is na\"ively combined, ambiguities could easily arise. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
Without strong modular reasoning principles to rely upon, it is difficult to construct large software systems and ecosystems. 
%This means that it is difficult to construct large software systems and ecosystems from components written in different dialects. 
The only reasonable approach for software engineers who must program ``in the large'' is to eschew constructs housed in dialects and settle on a single common language, where they can  rely on the guarantees provided by its type and module system. %The designers of these languages may slowly adopt the most generally useful new ideas. 
This has left many useful constructs for programming ``in the small'' obscure.% It is safe to say that in contemporary times, language dialects should be considered harmful. % It is little wonder that many programmers eschew esoteric language dialects in favor of languages that stopped evolving decades ago.

\section{Contributions}
Our aim in this thesis is to make dialect formation less frequently necessary by designing language mechanisms that give library providers the ability to express derived syntactic and semantic constructs from within the language, in a manner that ensures that these constructs can be reasoned about separately and composed arbitrarily.

%Such mechanisms are colloquially known as \emph{extension mechanisms} because the sorts of constructs they are used to express are those that would otherwise require extensions to the language itself. %Defining a new construct then corresponds to the usual notion of language extension. We will make this intuition more precise later.

%in languages without such mechanisms, expressing such constructs would require extensions to the language itself (a better qualifier might thus be \emph{extension-mitigating}). %We will make these notions more precise as we continue. %Thus, these mechanisms precisely capture the notions of language extension and language composition.

%The mechanisms we propose are designed to decrease the number of constructs that need to be built in \emph{a priori} by a language designer. However, there are still some basic syntactic and semantic decisions that must unavoidably be made.

To provide a coherent vehicle for these mechanisms, we will introduce a new general-purpose language called Wyvern. Let us briefly summarize its organization. Syntactically, Wyvern has a layout-sensitive textual concrete syntax (i.e. indentation is meaningful). This choice is not fundamental to our proposed mechanisms, but it will be useful for a class of examples that we will discuss later.  Semantically, Wyvern consists of a \emph{module language} and a \emph{core language}. The module language is based directly on the Standard ML module language, with which we assume familiarity for the purposes of this proposal \cite{harper1997programming}. The core language (i.e. the language of types and expressions) is split into an \emph{external language} (EL) and an \emph{internal language} (IL). The semantics of EL constructs are specified by a type-directed translation to the much simpler IL. Notionally, this can be thought of as simply a shifting of the first stage of a type-directed compiler into the language specification \cite{tarditi+:til-OLD}. The internal language we will use is the polymorphic lambda calculus with binary product and sum types and recursive types (features like state and exceptions could also be included, but for simplicity, we will omit them) \cite{pfpl}. %Again, we will make these statements more precise as we continue.%The external language provides local type inference \cite{Pierce:2000:LTI:345099.345100}. % (for the purposes of this work).

We will begin by describing Wyvern's mechanisms for expressing new forms of derived concrete syntax, assuming for the moment that the EL is otherwise specified in a standard way, in Section \ref{sec:syntax}. We then turn in Section \ref{sec:semantics} to a mechanism for introducing new semantic constructs -- in particular, new base types and operators -- into the EL. Constructs that are built in to languages like ML, like $n$-ary tuples/records, labeled sums and others, can in Wyvern be realized as modes of use of this mechanism. In both cases, the mechanisms can be broadly understood as novel forms of \emph{static metaprogramming}. Users define each new syntactic and semantic construct by writing code that generates terms and types. This code is  evaluated statically, in the course of typechecking EL terms.

%Languages that are organized around mechanisms like these are colloquially known as \emph{extensible languages}.

For each mechanism that we introduce, we will first demonstrate its expressive power by showing how constructs that are, or would need to be, built directly in to contemporary languages can in Wyvern be expressed in libraries. To demonstrate that these mechanisms are theoretically sound, we will then develop a formal specification and proofs of various important metatheoretic properties. The most interesting of these will be various \emph{modular reasoning principles} that guarantee that user-defined syntactic and semantic constructs can be reasoned about separately and then used together in any combination. These justify our classification of Wyvern as a \emph{modularly extensible programming language}.

%Designing a modularly extensible language is a challenge because giving away too much control can easily make reasoning about program behavior difficult and lead to problems that arise only when extensions are used at the same time. For example, a library provider given free reign over syntax and semantics could introduce parsing conflicts, violate {type safety}, and  interfere with other extensions. We aim to design mechanisms that eliminate these possibilities, while providing a degree of expressive power sufficient to express a variety of constructs that are, or would need to be, built directly in to today's languages. %ll constructs that can be defined using a mechanism we will introduce can be reasoned about separately and composed arbitrarily.% With these mechanisms, we anticipate dialect formation being far less common. %Indeed, all of these problems have plagued previous  designs.

% Common dialects of ML that suffice for \cite{harper1997programming}. %We will .

   %Our challenge will be to design mechanisms that achieve these goals while also being powerful enough to express features that today require dialects.%Our challenge will be to design such extension mechanisms, while still retaining significant expressive power -- we wish to permit library-based expression of a range of features that today are, or would need to be, built into a system by its designers.%It can also make it more difficult to define new sorts of tools for working with the language because they cannot rely on there being a fixed abstract syntax (the canonical \emph{expression problem}). For programmers, 
%It can make it difficult to understand and reason about the type of an unfamiliar term, a critical facet of program comprehension (the \emph{typing discipline problem}).

%So our goal will be to design extension mechanisms that handle these metatheoretic and composition-related issues while still  permitting library-based expression of a range of features that today are, or would need to be, built into a system by its designers. %More specifically, we will show how to delegate control, in a methodical manner, to \emph{static functions}, i.e. user-defined functions written in a \emph{static language}. By layering these mechanisms over a fixed typed internal language, constraining the static language appropriately and validating the code it is used to generate, we will retain key  metatheoretic properties and arrive at powerful modular reasoning principles. %By associating these static functions with type constructors, forming \emph{active type constructors},  we will further show how they can enjoy a usage profile essentially identical to that of built in features. % while maintaining key metatheoretic properties and modular reasoning principles. % By keeping the abstract syntax fixed and layering these mechanisms atop a fixed typed internal language (essentially lifting the first stages of compilation into the language), we avoid most aspects of the expression problem. %By delegating control to user-defined functions provided alongside user-defined type constructors (forming \emph{active type constructors}), we can give library providers substantially more control over these features of the system. 
%We structure each mechanism that we introduce as a \emph{bidirectional type system} combined with an  \emph{elaboration semantics} targeting a fixed typed internal language. 
%By constraining these functions and enforcing critical abstraction barriers between extensions, we will ensure key metatheoretic properties of the language and system components and guarantee that extensions can be reasoned about modularly. %We call user-defined types that introduce new features into the system in this way \emph{active types}.  % also highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.

%The result of this work will be the design of a \emph{modularly extensible programming language} called Wyvern.   and end with a brief foray into editor services in Sec. \ref{sec:editor-services}.\todo{fix sections}

\section{Modular Derived Syntax}\label{sec:syntax}
Most contemporary computer programming languages specify a textual concrete syntax. Because it serves as the language's human-facing user interface, it is common practice to include  derived syntactic forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally. % (i.e. considering cognitive dimensions \cite{green1996usability}). 
For example,  list syntax is built in to most dialects of ML, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, a programmer can equivalently write \lstinline{[1, 2, 3]}. Many languages go further, building in syntax associated with other types of values, like vectors (SML), arrays (Ocaml), commands (Haskell) and syntax trees (Scala).  %The desugaring from the latter to the former is specified by the language itself. %Typically, the language designer controls what forms of derived syntax are built in to the language.

Rather than privileging  particular data structures, Wyvern instead exposes mechanisms that allow library providers to introduce new derived syntactic constructs on their own. %Lists need no special consideration from the language specification.
%The purpose of this section is to motivate and then introduce Wyvern's syntax extension mechanisms. %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
To motivate our desire for this level of syntactic control, we begin in Sec. \ref{sec:examples} with the example of regular expression patterns expressed using abstract types, showing how the usual workaround of using strings to introduce patterns is not ideal. We then survey existing syntax extension mechanisms in Sec. \ref{sec:syntax-existing}, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs. In Sec. \ref{sec:syntax-contributions}, we outline our proposed mechanisms and discuss how they resolve these issues. We conclude in Sec. \ref{sec:syntax-timeline} with a timeline for remaining work.

\subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
Let us begin by taking the perspective of a \emph{regular expression} library provider. Recall that regular expressions are a common way to capture patterns in strings \cite{Thompson:1968:PTR:363347.363387}. We will assume that the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Wyvern supports these as case types, which are analagous to datatypes in ML:

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-20pt}
\section{Motivation}\label{motivation}
% \begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
% and generality in language design is a serious unsolved 
% problem.} -- John Reynolds, 1970 \cite{Reynolds70}\end{quote}
%%One might exp ``generality''
%Given a well-designed general-purpose programming language, programmers should be able to express the constructs that they need in libraries, as modes of use of existing constructs. New language dialects should be needed exceptionally rarely.

Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
There has been substantial progress on many of the foundational semantic issues that are relevant to the design of general-purpose languages over the past  decades, much of it driven by advances in type theory. However, when more pragmatic language design issues are brought into consideration, a broadly agreeable set of primitives remains elusive, as evidenced by the diverse array of dialects that surround around all major con\-temporary languages. 
For example, let us consider the many dialects of ML. Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with sums, products and recursive types, they all introduce various syntactic and semantic extensions and tweaks of their own novel design (we discuss a number of examples below). Tools for  constructing new ``domain-specific'' language dialects also continue to proliferate. % The situation is similar across all major language lineages. 
Why are well-informed programmers and researchers so often tempted to construct  dialects of a language like ML? 
%Let us investigate why this is.
%What kinds of constructs are not expressible in languages like SML?

% Unlike libraries organized by a module system, language dialects are not trivially composable.%Constructing a software system using libraries written in different dialects is difficult because they cannot always be composed. %This suggests that library-based mechanisms are still not general enough to encompass many desirable modes of expression.   %Library implementations are not yet satisfying.

Perhaps the most common reason is simply that, for human {programmers}, the \emph{syntactic cost} of expressing a construct of interest using general-purpose primitives is not always ideal. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}. This gives us precisely the semantics that we need, but the syntactic cost  is not quite ideal. In situations like this, library providers are often tempted to construct a \emph{derived syntactic dialect}, i.e. a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to their library. For example, Ur/Web is a derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}. Indeed, nearly all major languages primitively build in derived syntax for constructs that are otherwise expressed in the standard library, e.g. derived list syntax in Standard ML (SML) \cite{harper1997programming,mthm97-for-dart}. Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, which we will return to later, help lower the engineering costs of constructing derived syntactic dialects, contributing to their proliferation.% contributing to their proliferation.

%Many other data structures can similarly benefit from a decrease in syntactic cost. For example, we  will consider {regular expression patterns} encoded using abstract data types in Sec. \ref{sec:syntax}. 
%The syntactic cost of  such encodings, considered in several ways, can be reduced by using a dialect that builds in {syntactic sugar}  for regular expression patterns, and there are many tools that make constructing such syntactic dialects simple, e.g. Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (though as we will see, they come with costs of their own).   %Put another way, such embeddings may not preserve the cognitive cost associated with a construct (qualitative criteria are generally used to establish this \cite{green1996usability}).

More advanced dialects introduce  new type structure, going beyond what is possible by a context-free elaboration to an existing language. For example, SML goes beyond the simple nullary and binary product types common in simpler calculi, instead exposing record types. Other dialects have explored ``record-like'' primitives that support functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
, mutable rows, methods, prototypic dispatch and various other embellishments. Ocaml builds in a more  sophisticated object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g. $\mathtt{sprintf}$ \cite{ocaml-manual}. ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.  ML5 builds in primitives for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}. Manticore builds in parallel arrays \cite{conf/popl/FluetRRSX07}. Alice ML builds in futures and promises \cite{AliceLookingGlass}. MLj builds in essentially the entire Java programming language, to support typesafe interoperation with existing Java code \cite{Benton:1999:IWW:317636.317791}. Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples. Tools like proof assistants and logical frameworks are designed to support specifying and reasoning about dialects like these, and tools like compiler generators and language workbenches lower the engineering cost of implementing them.

%We will also discuss labeled products, which unlike records maintain a row ordering (this makes it possible to introduce them without explicit labels, but eliminate them using labels).

@@

The problem with this practice of using language dialects as vehicles for new syntactic and semantic primitives is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect. %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,  
This would fundamentally require combining the two dialects into a single language. Left unconstrained, this is an ill-posed problem, i.e. given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them. Even when the dialects are specified using a formalism that does provide some way to na\"ively combine two languages, there is generally no guarantee that the composition will conserve important syntactic and semantic properties. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type. Though each is known to have an unambiguous concrete syntax in isolation, when their syntax is na\"ively  combined (e.g. by Camlp4), ambiguities arise.  Any putative ``combined language'' must formally be considered a  distinct system for which one must derive all metatheorems of interest anew, guided only informally by those derived for the dialects individually. % would generally attempt to combine their semantics, particularly when they are specified in very different ways. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
Due to this paucity of modular reasoning principles, language dialects are not practical artifacts for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

Dialects do, however, have an indirect impact on the practice of programming because the language designers that control comparatively popular languages, like Ocaml and Scala, can occasionally be convinced to incorporate  ideas from dialects  into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
These major languages have thus become more expressive over time. However, this has come at a cost: they have also ballooned in size. Moreover, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntactic primitives, our group has gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects. A conservative approach to incorporating new primitives is only sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below. Recalling the words of  Reynolds, which are clearly as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition. %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.

\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
and generality in language design is a serious unsolved 
problem.}%\begin{flushright}-- John Reynolds (1970)\end{flushright}
\end{quote}

\newpage
 There are two possible paths forward for the subset of the language design community interested in keeping general-purpose languages small and free of \emph{ad hoc} primitives. One, exemplified (arguably) by Standard ML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, these primitives trade away expressive power or have high syntactic cost. % that this implies.  
The other path forward is to press on in the search for even more general language primitives, i.e. primitives that are so expressive that they allow us to degrade a broad class of existing primitives, including those found in dialects, into modularly composable library constructs, where they can be evaluated on their individual merits by programmers. % This should render the construction of new dialects increasingly unnecessary. 
Such primitives do sometimes arise. For example, Ocaml recently introduced  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}. Using GADTs, Ocaml was able to move some machinery for typechecking operations that use format strings, like \texttt{printf}, out of the language and into the standard library (though some machinery remains built in). Our broad aim will be to introduce a new general-purpose programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes even more dramatic steps down this second path. %Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .

%Viewed ``dually'', one might equivalently ask for a language that builds in a core that is as small as possible, but provides expressive power comparable to languages with much larger cores. This is our goal in the work being proposed.

\section{Proposed Contributions}

Verse has a module system taken directly from SML, but its core language is organized in a novel manner around a minimal typed lambda calculus, called the Verse \emph{internal language}, introduced in Section \ref{sec:verse}. In lieu of typical {external} (i.e. programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs: \textbf{typed syntax macros} (TSMs),  in Section \ref{sec:syntax}, and \textbf{metamodules},  in Section \ref{sec:metamodules}. Briefly summarized, TSMs subsume the need for derived syntax specific to library constructs (e.g. list syntax in SML) by giving library providers static control over the parsing and elaboration of delimited segments of textual concrete syntax. Metamodules give library providers static hooks directly into the semantics of the Verse EL, allowing  library providers to, for example, define the type structure of records (and various ``record-like'' constructs, as discussed above) by type-directed translation to the internal language (which builds in only nullary and binary products). Both TSMs and metamodules can be understood as novel \emph{metaprogramming} primitives, because they involve static code generation. We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.

The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of the language that have typically been under the monolithic control of the language designer. If we are not careful, many of the problems mentioned above as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer. The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended. We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner. In particular, syntactic ambiguities will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles. Library providers will be able to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We will assume that simple naming conflicts can be avoided extrinsically in some suitable manner, e.g. by using a URI-based naming scheme as in the Java ecosystem.}

\subsection{Thesis Statement}
In summary, we propose a thesis defending the following statement:
\begin{quote}
A programming language can  give library providers the power to meta\-pro\-gram\-matic\-ally express new derived syntax and external type structure atop a minimal typed internal language, using  primitives that maintain a hygienic type discipline and modular reasoning principles. These  primitives are  expressive enough to subsume the need for a variety of primitives that are, or would need to be, built in to comparable contemporary languages.
\end{quote}

\subsection{Disclaimers}
Before we continue, it may be useful to explicitly acknowledge that completely eliminating the need for dialects would indeed be asking for too much: certain design decisions are fundamentally incompatible with others or require coordination across a language design. We aim only to decrease the need for dialects.% out a larger design space within a single language, Verse.%a subset of constructs that can be specified by a semantics of a certain ``shape'' specified by Verse (we will make this more specific later). %There is nothing ``universal'' about Verse.

It may also be useful to explicitly acknowledge that library providers could leverage the primitives we introduce   to define constructs that are in rather poor taste. We  expect that in practice, Verse will come with a standard library defining a carefully curated collection of standard constructs, as well as guidelines for advanced users regarding when it would be sensible to use the primitives we introduce (following the example of languages that support operator overloading or typeclasses, which also have the potential for ``abuse''). %The vast majority of programmers should not use the primitives that we introduce directly.

Finally, we are not interested here in languages that feature full-spectrum dependent types, which blur the phase separation between compile-time and run-time, though we conjecture that the primitives we introduce could be introduced into languages like Gallina (the ``external language'' of the Coq proof assistant) with some modifications. Verse should be compared to languages that maintain a phase separation, like ML and Scala. %Our interest is in making sure that the standard library is no more privileged than any other library. % Our interest is not in creating a universal language, only one that is as expressive as reasonably possible. %only in a subset of all constructs that can be specified in a mutually orthogonal manner by a semantics of a certain prototypic ``shape''. We will make this more specific later.

\section{Verse}\label{sec:verse}
Let us begin by briefly summarizing how Verse is organized and specified. Verse consists of a \emph{module language} and a \emph{core language}. The module language is based directly on SML's module language,  which we assume familiarity with for the purposes of this work \cite{harper1997programming,MacQueen:1984:MSM:800055.802036}. We will give an example of its use in Section \ref{sec:syntax}, but do not plan to specify it in detail.

The core language is organized much like the first stage of a type-directed compiler (e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}), consisting of a user-facing \emph{typed external language} (EL) specified by type-directed translation to a minimal \emph{typed internal language} (IL). The main judgements in the specification of the EL take the following form (we omit mention of various contexts for now):
\\[1ex]
$\begin{array}{ll}
\textbf{Judgement} & \textbf{Pronunciation}\\
e \Rightarrow \sigma \leadsto \iota & \text{External expression $e$ synthesizes external type $\sigma$ and has translation $\iota$.}\\
e \Leftarrow \sigma \leadsto \iota & \text{External expression $e$ analyzes against external type $\sigma$ and has translation $\iota$.}\\
\sigma~\mathtt{type} \leadsto \tau & \text{Static expression $\sigma$ is an external type with translation $\tau$.}
\end{array}
$\\[1ex]
Note that the typing judgements are \emph{bidirectional}, i.e. there is a distinction between \emph{type synthesis} (the type is an ``output'') and \emph{type analysis} (the type is an ``input'') \cite{Pierce:2000:LTI:345099.345100}. This is to  allow us to explicitly specify how Verse's \emph{local type inference} works, and in particular, how it interacts with the primitives that we will introduce. Like Scala, we intentionally do not aim to support global type inference, both because it is not compatible with the primitives we introduce as described and because we believe that in practice, the cognitive cost of the error messages that arise outweigh its marginal utility relative to local type inference. Note also that, for concision, we will write ``expression'' and ``type'' without qualification to refer to  external expressions and types, respectively.

% The reason why we refer to the syntactic sort $\sigma$ as a ``static expression'' will become more clear later.

%The main novelty of our work is that the Verse EL builds in very little syntax or type structure \emph{a priori}. %Instead, \textbf{metamodules} and \textbf{typed syntax macros} (TSMs) are expressive enough to render this unnecessary. Although metamodules are in some sense prior to TSMs, we will describe TSMs first, temporarily assuming that the EL does build in standard type structure, for pedagogical purposes.

Our design is largely insensitive to the details of the IL. The only strict requirements are that the IL is type safe and supports parametric type abstraction. For our purposes, we will keep things simple by using the strictly evaluated polymorphic lambda calculus with nullary and binary product and sum types and recursive types, which we assume the reader has familiarity with and for which all the necessary metatheorems are well-established (we follow \cite{pfpl}). The main judgements in the static semantics of the IL (again omitting contexts for now) thus take the following form:
\\[1ex]
$
\begin{array}{ll}
\textbf{Judgement} & \textbf{Pronunciation}\\
\iota : \tau & \text{Internal expression $\iota$ has internal type $\tau$.}\\
\tau~\mathtt{type} & \text{Internal type $\tau$ is valid.}
\end{array}
$\\

\noindent
The dynamic semantics can be specified as a transition system with  judgements of the following form:
\\[1ex]
$
\begin{array}{ll}
\textbf{Judgement} & \textbf{Pronunciation}\\
\iota \mapsto \iota' & \text{Internal expression $\iota$ transitions to $\iota'$.}\\
\iota~\mathtt{val} & \text{Internal expression $\iota$ is a value.}
\end{array}
$
\\[1ex]
The iterated transition judgement $\iota \mapsto^{*} \iota'$ is the reflexive, transitive closure of the transition judgement, and the evaluation judgement $\iota \Downarrow \iota'$ is derivable iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$.

Primitive features like state, exceptions, concurrency primitives, scalars, arrays and others characteristic of a first-stage compiler IL would also be included in practice, and in some cases this would affect the shape of the internal semantics, but we omit them when their inclusion would not meaningfully affect our discussion. %Features like record types and labeled sum types, on the other hand, are not built in.

\section{Modularly Metaprogrammable Textual Syntax}\label{sec:syntax}
Verse, like most major contemporary programming languages, specifies a textual concrete syntax. We have chosen to specify a layout-sensitive textual concrete syntax (i.e. newlines and indentation are not ignored). This design choice is not  fundamental to our proposed contributions, but it will be useful for cleanly expressing a class of examples that we will discuss later. We will specify key aspects of its concrete syntax with an \emph{Adams grammar} \cite{Adams:2013:PPI:2429069.2429129} (such a specification for Wyvern, which has a very similar syntax, can be found in \cite{TSLs}), but for the purposes of this proposal, we will simply introduce Verse's concrete syntax by example as we go on.

A programming language's concrete syntax serves as its human-facing user interface, so it is common practice to build in  derived forms (colloquially, \emph{syntactic sugar}) that capture common idioms more concisely or naturally. % (i.e. considering cognitive dimensions \cite{green1996usability}). 
For example, derived list syntax is built in to most dialects of ML, so that instead of having to write \lstinline{Cons(1, Cons(2, Cons(3, Nil)))}, a programmer can equivalently write \lstinline{[1, 2, 3]}. Many languages go further, building in syntax associated with various other types of data, like vectors (SML/NJ), arrays (Ocaml), commands (Haskell), syntax trees (Scala), XML data (Scala, Ur/Web) and SQL queries (F\#, Ur/Web). %The desugaring from the latter to the former is specified by the language itself. %Typically, the language designer controls what forms of derived syntax are built in to the language.

Rather than privileging  particular library constructs with primitive syntactic support, Verse exposes primitives that allow library providers to introduce new derived syntax on their own. %Lists need no special consideration from the language specification.
%The purpose of this section is to motivate and then introduce Verse's syntax extension mechanisms. %For forms with a clear analogy to a form in Standard ML, we will assume the  semantics are analagous without providing details.
To motivate our desire for this level of syntactic control, we begin in Sec. \ref{sec:examples} with a representative example: regular expression patterns expressed using abstract data types. We show how the usual workaround of using strings to introduce patterns is not ideal. We then survey existing alternatives in Sec. \ref{sec:syntax-existing}, finding that they involve an unacceptable loss of modularity and other undesirable trade-offs. In Sec. \ref{sec:syntax-contributions}, we outline our proposed mechanisms and discuss how they resolve these issues. We conclude in Sec. \ref{sec:syntax-timeline} with a timeline for remaining work.

\subsection{Motivating Example: Regular Expression Syntax}\label{sec:examples}
Let us begin by taking the perspective of a \emph{regular expression} library provider. Recall that regular expressions are a common way to capture patterns in strings \cite{Thompson:1968:PTR:363347.363387}. We will assume that the abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
The most direct way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as case types, which are analagous to datatypes in ML (we will see how the type structure  of case types can in fact be expressed in libraries later):
